Supplement to: Inducing Domain Specific Languages for Bayesian Program Learning

### 2018-02-07 paper.tex
progress on describing baseline comparisons

Our goal  is to induce a DSL and find good programs solving each of the tasks.
Our strategy is to iterate through three steps: (1) searching for programs that
solve the tasks, (2) learning a better neural recognition model -- which we use
to accelerate the search over programs -- and (3) improving the DSL.
The key observation here is that each of these three steps can bootstrap off each other:

\begin{description}
\item[ Searching for programs:]  Our program search  is informed by both the DSL and the recognition model.
When these improve we find better programs.
  \item[ Learning a recognition model:] The recognition model is trained both on samples from the DSL and on programs found by the search procedure. As the DSL improves and as search finds more programs, the recognition model gets both more data to train on and better data.
  \item[Improving the DSL:] We induce a DSL from the programs we have found so far which solve the tasks;
  as we solve more tasks, we can hone in on richer DSLs that more closely match the domain.
\end{description}

The Helmholtz machine~\cite{dayan1995helmholtz} introduced the idea of alternating between
learning a generative model and training a recognition model on samples from the generative model.
We apply this idea to program induction, motivating why we call our algorithm the \system.
Figure~\ref{feeding} diagrams how these three steps feed off of each other.
\begin{figure}\centering
  \begin{tikzpicture}
    \begin{scope}[shift = {(1,-1)}]
    \node[align = center](synthesis) at (6,4) {Search for \\programs: $p$};
    \node[align = center](DSL) at (3,1) {DSL: $\mathcal{D}$};
    \node[align = center](recognitionModel) at (9,1) {Recognition \\model: $q$};

@@

    \draw [->,thick,dashed] (DSL.north) to[out = 90,in = 180] node[fill=white,align = center]{  \footnotesize{Provides}\\ \footnotesize{inductive bias}} (synthesis.west);
    \draw [->,thick,dashed] (recognitionModel.north) to[out = 90,in = 0] node[fill=white]{{\footnotesize Makes tractable}} (synthesis.east);
  \end{scope}
    \end{tikzpicture}
\caption{Our algorithm solves for programs, the DSL, and a recognition model. Each of these steps bootstrap off of the others.}  \label{feeding}
  \end{figure}

@@

\textbf{Neural recognition model: tractably maxing $\lowerBound$ w.r.t. the
  frontiers.}  Here we train a neural network, $q$, to predict a
distribution over programs conditioned on a task. The objective of $q$
is to assign high probability to programs $p$ where
$\probability[x,p|\mathcal{D},\theta]$ is large.  With $q$ in hand we can find programs for
the frontier $\mathcal{F}_x$ by searching for programs that maximize
$q(p|x)$.
The network $q$ exploits the structure of the DSL $\mathcal{D}$:
rather than directly predicting a distribution over $p$ conditioned on $x$,
it predicts a weight vector, $\theta^{(x)}$, and we define $q(p|x)\triangleq \probability[p|\mathcal{D},\theta = q(x)]$.
This taps into the intuition that programming is primarily a top-down activity:
as human programmers, we first decide what kind of
programming constructs we might need to use,
and then we figure out how to assemble them in to the desired program.
%% This approach implements an amortized
%% inference scheme~\cite{le2016inference,ritchie2016deep} for the generative model in
%% Figure~\ref{graphicalModel}.

@@

Why enumerate, when the program synthesis community has invented many
sophisticated algorithms that search for programs?~\cite{solar2008program,schkufza2013stochastic,feser2015synthesizing,osera2015type,polozov2015flashmeta}.
We have two reasons:% for using the simple strategy of enumerating programs:
\begin{itemize}
\item A key point of our work is that learning the DSL, along with a neural recognition model, can make program induction tractable, even if the search algorithm is very simple.
\item Enumeration is a general approach that can be applied to any program induction problem. Many of these more sophisticated approaches require special conditions on
  the space of of programs.
\end{itemize}
A main drawback of an enumerative search algorithm is that we have no
efficient means of solving for arbitrary constants that might occur in the
program. In Section~\ref{regressionSection},
we will show how to find programs with real-valued constants
by automatically differentiating through the program and setting the constants using gradient descent.
%% In Section~\ref{textSection}
%% we will show that the bottom-up neural recognition model can learn
%% which discrete constants should be included in a program.

@@

\subsection{Boolean circuits}
As a toy domain, we consider the problem of learning
to build the Boolean circuits out of logic gates.
Now, if we equipped our agent with a full repertoire of
circuit primitives -- XOR gates, multiplexers, etc. -- this problem would be easy,
and we could use sophisticated symbolic search algorithms like SMT solvers to synthesize circuits.
Instead, we gave \system only a NAND gate,
and tasked it with building a set of 500 random circuits made out of
AND, NOT, and OR gates,
as well as 2 or 4 input multiplexers.
This experimental setup is an extension of one introduced in the original EC algorithm~\cite{Dechter:2013:BLV:2540128.2540316}.
For this domain our recognition model is a simple multilayer perceptron
whose inputs are the input/output examples.
Figure~\ref{circuitLearningCurve} shows our algorithm learning to build circuits; Table~\ref{circuitExperimentTable}

\begin{figure}
  \includegraphics[width = \columnwidth]{figures/circuit.png}
  \caption{\system learning to build Boolean circuits. We start out with only NAND gates and learn DSL that contains ...}\label{circuitLearningCurve}
\end{figure}

@@

\begin{figure}\centering
  \includegraphics[width = 7cm]{figures/symbolicRegression.png}
  \caption{DSL learning curve for symbolic regression.}\label{regressionLearningCurve}
\end{figure}

@@

We automatically generated  600 string editing problems,
along with input/output examples (Figure~\ref{exampleTextProblem}).

@@

  \centering
  \begin{tabular}{c}
    String editing primitives\\
    \begin{tabular}{l}
    \toprule
    \texttt{0}, \texttt{+1},\texttt{-1}\\
    \texttt{split}, \texttt{join}, \texttt{map}, \texttt{length}, \texttt{slice}\\
    \texttt{append}, \texttt{chr->str}, \texttt{trim}, \texttt{nth}\\
    \texttt{lowercase}, \texttt{uppercase}, \texttt{capitalize}\\
    \texttt{""},    \texttt{","},     \texttt{" "},    \texttt{"<"}, \texttt{">"}  
  \\\bottomrule
    \end{tabular}\\
\\    Subset of learned DSL\\
    \begin{tabular}{l}
      \toprule
      $f_0($\texttt{c}$) = $\texttt{(+ (chr->str c))}\\
      $f_1($\texttt{i,c,s}$) = $      \texttt{(+ (nth i (split c s))}\\\hspace{2.55cm}\texttt{ (chr->str c))}\\
      $f_2($\texttt{s,c}$) = $\texttt{((}$f_0$\texttt{ c) (}$f_1$\texttt{ (1+ 0) c s))} \\
      \hspace{1cm}($f_2$: \emph{Extract substring delimited by character})
    %% $f_0($\texttt{i,c,s}$) = $    \texttt{(+ (index i (split c s)) (chr->str c))}\\
    %% $f_1 = $\texttt{(join "")}
  \\\bottomrule
    \end{tabular}
  \end{tabular}
  \caption{Top: provided string editing primitives. Bottom: some induced DSL primitives.}\label{textPrimitives}
  \end{table}
\begin{figure}

  \includegraphics[width = 7cm]{figures/text.png}
  \caption{Learning curves for string editing programs}\label{textLearningCurve}

\end{figure}

>>>>

Our goal  is to induce a DSL and find good programs solving each of the tasks.
Our strategy is to iterate through three steps: (1) a \emph{wake} phase where we search for programs, and then a pair of \emph{sleep} phases: (2) learning a neural recognition model -- which we use
to accelerate the search over programs -- and (3) improving the DSL.
The key observation here is that these three steps bootstrap off each other:

\noindent \textbf{Searching for programs:}  Our program search  is informed by both the DSL and the recognition model.
When these improve we find better programs.

\noindent\textbf{Learning a recognition model:} The recognition model is trained both on samples from the DSL and on programs found by the search procedure. As the DSL improves and as search finds more programs, the recognition model gets both more data to train on and better data.

\noindent\textbf{Improving the DSL:} We induce a DSL from the programs we have found so far which solve the tasks;
  as we solve more tasks, we can hone in on richer DSLs that more closely match the domain.

%% The Helmholtz machine~\cite{dayan1995helmholtz} introduced the idea of alternating between
%% learning a generative model and training a recognition model on samples from the generative model.
%% We apply this idea to program induction, motivating why we call our algorithm the \system.
Figure~\ref{feeding} diagrams how these three steps feed off of each other.
\begin{figure}\centering
  \begin{tikzpicture}
    \begin{scope}[shift = {(1,-1)}]
    \node[align = center](synthesis) at (6,4) {Search for \\programs: $p$};
    \node[align = center](DSL) at (3,1) {DSL: $\mathcal{D}$};
    \node[align = center](recognitionModel) at (9,1) {Recognition \\model: $q$};

@@

    \draw [->,thick,dashed] (DSL.north) to[out = 90,in = 180] node[fill=white,align = center]{  \footnotesize{Provides}\\ \footnotesize{inductive bias}} (synthesis.west);
    \draw [->,thick,dashed] (recognitionModel.north) to[out = 90,in = 0] node[fill=white]{{\footnotesize Makes tractable}} (synthesis.east);
  \end{scope}
    \end{tikzpicture}
\caption{Our algorithm solves for programs, the DSL, and a recognition model. Each of these steps bootstrap off of the others in a Helmholtz-machine inspired wake/sleep inference algorithm.}  \label{feeding}
  \end{figure}

@@

\textbf{Neural recognition model: tractably maxing $\lowerBound$ w.r.t. the
  frontiers.}  Here we train a neural network, $q$, to predict a
distribution over programs conditioned on a task. The objective of $q$
is to assign high probability to programs $p$ where
$\probability[x,p|\mathcal{D},\theta]$ is large, because including those programs
in the frontiers will most increase $\lowerBound$.  With $q$ in hand we can find programs for
the frontier $\mathcal{F}_x$ by searching for programs that maximize
$q(p|x)$.
The network $q$ exploits the structure of the DSL $\mathcal{D}$:
rather than directly predicting a distribution over $p$ conditioned on $x$,
it predicts a weight vector, $\theta^{(x)}$, and we define $q(p|x)\triangleq \probability[p|\mathcal{D},\theta = q(x)]$.
This taps into the intuition that programming is primarily a top-down activity:
as human programmers, we first decide what kind of
programming constructs we might need to use,
and then we figure out how to assemble them in to the desired program.
%% This approach implements an amortized
%% inference scheme~\cite{le2016inference,ritchie2016deep} for the generative model in
%% Figure~\ref{graphicalModel}.

@@

Why enumerate, when the program synthesis community has invented many
sophisticated algorithms that search for programs?~\cite{solar2008program,schkufza2013stochastic,feser2015synthesizing,osera2015type,polozov2015flashmeta}.
We have two reasons:
(1) A key point of our work is that learning the DSL, along with a neural recognition model, can make program induction tractable, even if the search algorithm is very simple.
(2) Enumeration is a general approach that can be applied to any program induction problem. Many of these more sophisticated approaches require special conditions on
  the space of of programs.

A main drawback of an enumerative search algorithm is that we have no
efficient means of solving for arbitrary constants that might occur in the
program. In Section~\ref{regressionSection},
we will show how to find programs with real-valued constants
by automatically differentiating through the program and setting the constants using gradient descent.
%% In Section~\ref{textSection}
%% we will show that the bottom-up neural recognition model can learn
%% which discrete constants should be included in a program.

@@

\subsection{Boolean circuits}
As a toy domain, we consider the problem of learning
to build the Boolean circuits out of logic gates.
Now, if we equipped our agent with a full repertoire of
circuit primitives -- XOR gates, multiplexers, etc. -- this problem would be easy,
and we could use sophisticated symbolic search algorithms like SMT solvers to synthesize circuits.
Instead, we gave \system only a NAND gate,
and tasked it with building a set of 500 random circuits made out of
AND, NOT, and OR gates,
as well as 2 or 4 input multiplexers.
This experimental setup is an extension of one introduced in the EC algorithm~\cite{Dechter:2013:BLV:2540128.2540316}.
For this domain our recognition model is a simple multilayer perceptron
whose inputs are the input/output examples.
Figure~\ref{circuitLearningCurve} shows our algorithm learning to build circuits; Table~\ref{circuitExperimentTable}

%% \begin{figure}
%%   \includegraphics[width = \columnwidth]{figures/circuit.png}
%%   \caption{\system learning to build Boolean circuits. We start out with only NAND gates and learn DSL that contains ...}\label{circuitLearningCurve}
%% \end{figure}

@@

The neural recognition model accelerates inference,
but given enough iterations of our algorithm,
a lesioned version without the recognition model eventually comes close to catching up (red curves in Figure~\ref{regressionLearningCurve}; No NN in Table~\ref{regressionComparison}).
We compare our model on held-out tasks against 4 baselines (Table~\ref{regressionComparison}):
No NN, which lesions $q$;
RF/DC,
which holds the generative model $(\mathcal{D},\theta)$ fixed and learns
a $q$ only from samples from the fixed generative model --
this is how RobustFill and DeepCoder are trained;
PCFG, which lesions $q$, fixes $\mathcal{D}$, and learns only $\theta$ (so it is analogous to learning the parameters of a PCFG);
and Enum, which does no learning and just enumerates a large frontier.
\begin{figure}\centering
  \includegraphics[width = 7cm]{figures/regression.eps}
  \caption{Learning curves for symbolic regression. Green: \system. Red: no neural network. Notice how the model without the neural network eventually almost catches up.}\label{regressionLearningCurve}
\end{figure}

\begin{table}
\begin{tabular}{lllll}
  \toprule \system&No NN  & RF/DC & PCFG & Enum\\\midrule
  98\% solved&94\% \\
  $2.71\pm 0.07$ sec& $2.84\pm0.07$
  \\\bottomrule
\end{tabular}\caption{\% held out tasks solved and average solve time for symbolic regression}\label{regressionComparison}
\end{table}

@@

We automatically generated  600 string editing along with input/output examples (Figure~\ref{exampleTextProblem}), split into a 75/25 train/test split.

\begin{table}
\begin{tabular}{lllll}
  \toprule \system&No NN  & RF/DC & PCFG & Enum\\\midrule
  99\% solved\\
  $0.88\pm 0.04$ sec
  \\\bottomrule
\end{tabular}\caption{\% held out tasks solved and average solve time for string edit problems}
\end{table}

@@

  \centering
  \begin{tabular}{c}
    String editing primitives\\
    \begin{tabular}{l}
    \toprule
    \texttt{0}, \texttt{+1},\texttt{-1}\\
    \texttt{split}, \texttt{join}, \texttt{map}, \texttt{length}, \texttt{slice}\\
    \texttt{append}, \texttt{chr->str}, \texttt{trim}, \texttt{nth}\\
    \texttt{lowercase}, \texttt{uppercase}, \texttt{capitalize}\\
    \texttt{""},    \texttt{","},     \texttt{" "},    \texttt{"<"}, \texttt{">"}  
  \\\bottomrule
    \end{tabular}\\
\\    Subset of learned DSL\\
    \begin{tabular}{l}
      \toprule
      $f_0($\texttt{c}$) = $\texttt{(+ (chr->str c))}\\
      $f_1($\texttt{i,c,s}$) = $      \texttt{(+ (nth i (split c s))}\\\hspace{2.55cm}\texttt{ (chr->str c))}\\
      $f_2($\texttt{s,c}$) = $\texttt{((}$f_0$\texttt{ c) (}$f_1$\texttt{ (1+ 0) c s))} \\
      \hspace{1cm}($f_2$: \emph{Extract substring delimited by character})
    %% $f_0($\texttt{i,c,s}$) = $    \texttt{(+ (index i (split c s)) (chr->str c))}\\
    %% $f_1 = $\texttt{(join "")}
  \\\bottomrule
    \end{tabular}
  \end{tabular}
  \caption{Top: provided string editing primitives. Bottom: some induced DSL primitives.}\label{textPrimitives}
  \end{table}
%% \begin{figure}

%%   \includegraphics[width = 7cm]{figures/text.png}
%%   \caption{Learning curves for string editing programs}\label{textLearningCurve}

%% \end{figure}

### 2018-02-07 supplement.tex
more improvements to the introduction
progress on describing baseline comparisons

However, many $\lambda$-calculus expressions correspond to ill-typed programs, such as the program that takes the logarithm of the Boolean \texttt{true} (i.e., \texttt{log true}) or which applies the number five to the identity function
(i.e., $5 \; (\lambda x.x)$).
We use a well-established typing system for $\lambda$-calculus called \emph{Hindley-Milner typing}~\cite{pierce}, which is used in programming languages like OCaml.
The purpose of the typing system is to ensure that our programs never call a function with a type it is not expecting (like trying to take the logarithm of \texttt{true}).
Hindley-Milner has two important features:
Feature 1: It supports \emph{parametric polymorphism}: meaning that types can have variables in them, called \emph{type variables}. Lowercase Greek letters are conventionally used for  type variables.
For example, the type of the identity function is $\alpha\to\alpha$, meaning it takes something of type $\alpha$ and return something of type $\alpha$. A function that returns the first element of a list has the type $\texttt{list}(\alpha)\to\alpha$. Type variables are not the same has variables introduced by $\lambda$-abstractions.
Feature 2: Remarkably, there is a  simple algorithm for automatically inferring the polymorphic Hindley-Milner type of a $\lambda$-calculus expression~\cite{damas1982principal}.
A detailed exposition of Hindley-Milner is beyond the scope of this work.

@@

@@

Now what we are actually after is the parameters that maximize the joint log probability of the data+parameters, which I will write $J$:
\begin{align}
  J& = L + \log \text{D}(\theta|\alpha)\\
  &\stackrel{+}{\geq } \sum_x c(x)\log \theta_x - N \log \sum_x a(x)\theta_x  + \sum_x(\alpha_x - 1)\log \theta_x\\
  & = \sum_x (c(x) + \alpha_x - 1)\log \theta_x -  N \log \sum_x a(x)\theta_x
\end{align}
So you add the pseudocounts to the \emph{counts} ($c(x)$), but not to the \emph{possible counts} ($a(x)$).

>>>>

However, many $\lambda$-calculus expressions correspond to ill-typed programs, such as the program that takes the logarithm of the Boolean \texttt{true} (i.e., \texttt{log true}) or which applies the number five to the identity function
(i.e., $5 \; (\lambda x.x)$).
We use a well-established typing system for $\lambda$-calculus called \emph{Hindley-Milner typing}~\cite{pierce}, which is used in programming languages like OCaml.
The purpose of the typing system is to ensure that our programs never call a function with a type it is not expecting (like trying to take the logarithm of \texttt{true}).
Hindley-Milner has two important features:
Feature 1: It supports \emph{parametric polymorphism}: meaning that types can have variables in them, called \emph{type variables}. Lowercase Greek letters are conventionally used for  type variables.
For example, the type of the identity function is $\alpha\to\alpha$, meaning it takes something of type $\alpha$ and return something of type $\alpha$. A function that returns the first element of a list has the type $\texttt{list}(\alpha)\to\alpha$. Type variables are not the same has variables introduced by $\lambda$-abstractions.
Feature 2: Remarkably, there is a  simple algorithm for automatically inferring the polymorphic Hindley-Milner type of a $\lambda$-calculus expression~\cite{damas1982principal}.
Our generative model over programs performs Hindley-Milner type inference during sampling;
\texttt{unify} in the generative model uses the machinery of Hindley-Milner to
ensure that the generated programs have valid polymorphic types.
A satisfactory exposition of Hindley-Milner is beyond the scope of this paper,
but~\cite{pierce} offers a nice overview of lambda calculus and typing systems like Hindley-Milner.

@@

\section{Neural recognition model architecture}

The neural recognition model regresses from an observation (set of input/output pairs: $\left\{(i_n,o_n) \right\}_{n\leq N}$) to a $|\mathcal{D}| + 1$ dimensional vector. Each input/output pair is processed by an identical encoder network;
the outputs of the encoders are max pooled and passed to an MLP with 1 hidden layer, 16 hidden units, and a ReLU activation:
\begin{align}
  q(x) = \text{MLP}\left(\text{MaxPool}\left(\left\{\text{encoder}\left(i_n,o_n \right) \right\}_{n\leq N} \right) \right)
\end{align}
For the Boolean circuits and symbolic regression domains,
the inputs and outputs have a fixed dimensionality,
and the set of possible inputs is prespecified. For these fixed-dimensionality domains our encoder is an MLP with 1 hidden layer having 16  units and a ReLU activation function. The input to the encoder is
the sequence of program outputs $\left\{o_n \right\}_{n\leq N}$.

For the string editing and list domains,
the inputs and outputs are sequences. Our encoder for these domains is a bidirectional GRU with 16 hidden units that reads each input/output pair; we concatenate the input and output along with a special delimiter
symbol between them.
We MaxPool the final hidden unit activations in the GRU along both passes of the bidirectional GRU.

\section{Why not the ELBO bound?}

@@

Now what we are actually after is the parameters that maximize the joint log probability of the data and parameters $\theta$, which we  write $J$:
\begin{align}
  J& = L + \log \text{D}(\theta|\alpha)\\
  &\stackrel{+}{\geq } \sum_x c(x)\log \theta_x - N \log \sum_x a(x)\theta_x  + \sum_x(\alpha_x - 1)\log \theta_x\\
  & = \sum_x (c(x) + \alpha_x - 1)\log \theta_x -  N \log \sum_x a(x)\theta_x
\end{align}
We identify the parameters of the prior over $\theta$, $\alpha$, as  pseudocounts that are added to the \emph{counts} ($c(x)$), but not to the \emph{possible counts} ($a(x)$).
